---
title: "Calculation and interpretation of physiological indices derived from tree ring d13C"
output: html_notebook
---

Isotope Dendrochronology
Calculation and interpretation of physiological indices derived from tree ring d13C
COURSE: DISC, LTRR, UNIVERSITY OF ARIZONA. INSTRUCTORS: S. BELMECHERI & P. SZEJNER

The following code and comments should further your experience using tree ring and atmospheric d13C data to calculate WUE and other physiological indices. This exercise relies upon raw tree-ring d13C measurements, atmospheric d13C measurements, and uses the photosythesis biochemical model of Farqhuar 1982.

There are a few questions throughout to catalyze understanding and critical thinking. The analysis is broken into 4 main parts -including 20 questions, at the end you should have gone through many of the steps commonly used for reconstructing and interpreting iWUE derived from tree ring d13C. You will also be familiar with uncertainties related to such calculations and best practices to perform them.

You will need to download the R project fromchttps://github.com/SBelmecheri/DISCO-Ntebook3.git and to set your own working directory. Once you open the Notebook 3.Rproj in R studio  you can open the Notebook 3.Rmd file to proceed with the activity.

## PART 0: Setting up directories and files 

For the following exercises, you will need  the following files in one directory.

1. A file containing tree-ring d13C measurements. The filename is "LIL.txt"

2. A file containing atmospheric 13C data. The filename is "ATM.txt"

3. A file containing regional temperature data. The filename is "CRU_tt.txt"

You will need to change the directory in R so that this is the working directory.
The command "setwd" is used for this as in the following line:

```{r}
setwd("~/Documents/DISC/NOTEBOOK3")
```

You will also need to install and load the dplR and  rpmodel packages:
```{r}
library(dplR)
library(rpmodel)
```

You should see the above file names (and anything else in this directory) when you type (or copy and paste) the following line:
```{r}
list.files()
```

If you see these files listed, you're ready to roll!

## PART 1: Importing and plotting raw tree-ring isotope data [d13C] & Atmospheric Data 

Import and plot the raw tree ring d13C dataset:
```{r}
TR13 <- ts((read.table("LIL.txt", header=T)), start=1901, frequency=1)
ts.plot(TR13,ylab="d13C (permil.VPDB)",xlab="Years", col="#9C964A")
```
 
Import and subset atmospheric data: 
```{r}
ATM <- ts((read.table("ATM.txt", header=T)), start=1901, frequency=1)
# subset d13C
ATM.13C <- ATM[,"atm13C"]
# subset CO2
ATM.CO2 <- ATM[,"atmCO2"]
```

Plot the raw tree ring d13C  and the atmospheric d13C datasets:
```{r}
layout(matrix(1:2,nrow = 1, ncol = 2,byrow = T), widths = c(3,3),heights = c(3,3),respect = TRUE)
ts.plot(ATM.13C,ylab="d13C (permil.VPDB)",xlab="Years", col="#9C964A",main="d13C")
ts.plot(ATM.CO2,ylab="CO2 (ppm)",xlab="Years", main="CO2")

```
 
Question 1: Compare both trends. Describe and interpret the relationship between the decline in d13C and the increase in CO2 concentration. 

## PART 2: Calculating iWUE- SIMPLE APPROACH 

The Farquhar 1982 biochemical model describes the isotopic discrimination against 13C (Δ13C) during carbon diffusion and fixation by plants. The Farquhar equation can be written as follows:

Δ13C = a+(b-a)*(ci/ca)

a=4.4 is the fractionation due to CO2 diffusion in air through the stomata.

b= 28 is the apparent net fractionation by RuBisCO during carboxylation.

ci and ca are are the leaf intercellular and ambient partial pressure of CO2 (Pa), respectively.

From the equation above, ci can be calculated as follows:

ci =  ca*(Δ13C-a)/(b-a)

We know ca from measurements.ca is in ppm. You can use ca in ppm such as:
```{r}
# We know ca from measurements.ca is in ppm. you use use it as is 
ca <- ATM.CO2 # in ppm
```
Or use ca in Pa as follows:
```{r}
# you need  r package "rpmodel"
library("rpmodel")
```
and the function calc_patm {rpmodel} to calculate atmospheric pressure at sea level. The variable needed is elevation.
```{r}
elv= 1600 # Elevation above sea-level (m.a.s.l.)
patm<- calc_patm(elv, patm0 = 101325)
```
calculate ca as follows:
```{r}
ca <- ( 1.e-6 ) * ca * patm  # in Pa
```
convert ca to a time series:
```{r}
ca <- ts(ca, start=1901, frequency=1)
```

WARNING *********
Make sure you selected a ca in ppm or pa (by commenting the one you discard)
WARNING *********

Δ13C can then be calculated as follows:
```{r}
d13C.disc <- (ATM.13C-TR13)/(1+TR13/1000)
```
And ci as follows:
```{r}
ci <-  ca*(d13C.disc-a)/(b-a)
```
And iwue as follows:
```{r}
iwue <- (ca-ci)/1.6
```
plot the physiological indices:
```{r}
layout(matrix(1:6,nrow = 3, ncol = 2,byrow = F), widths = c(3,3,3),heights = c(3,3,3),respect = TRUE)

ts.plot(TR13,ylab="d13C (permil.VPDB)",xlab="Years", main="OBSERVATIONS")
ts.plot(d13C.disc,ylab="D13C (permil.VPDB)",xlab="Years", main="")
ts.plot(ca,ylab="CO2 (pa)",xlab="Years", main="")
ts.plot(ci,ylab="ci (pa)",xlab="Years", main="FARQUHAR-MODEL")
ts.plot(iwue,ylab=" iwue (umol/mol)",xlab="Years", main="")
ts.plot(ci/ca,ylab="ci/ca",xlab="Years", main="")

```
Question 2: Describe the trends of the various physiological indices.

Question 3: Based on what you have learned in the Isotope theory lectures, describe the trees'physiological response over time at this site.


## PART 3: Calculating iWUE- COMPREHENSIVE APPROACH 

In the Farquhar model, the d13C is the isotopic ratio for sugars fixed in the leaf. For tree-ring cellulose, a correction factor is used to account for the offset of d13C between leaf sugars and tree-ring cellulose. The offset results from post-photosynthetic fractionation processes (Gessler et al., 2014). 

The offset between whole wood and leaves is ~1.3 ± 0.2‰ for oak and conifer species. Additional isotopic offsets include the difference between tree-ring cellulose and bulk wood with an average value of ~1.3 ± 0.2‰, and isotopic depletion between primary assimilates and bulk leaf with values of -0.5 ± 1‰. 

Upscaling d13C measurements from tree rings to the leaf level can significantly improve estimates of Δ13C and hence reduce uncertainties in the determination of the ratio of ci to ca.

We can scale tree ring d13 to leaf level using a factor d.

d represents the sum of post-photosynthetic isotope fractionations between the leaf organic matter and the plant material considered. Choose the d value based on the studied material ( for this site, it is cellulose)

```{r}
d=1.9 # for d13C measured in wood
d=2.1 # for d13C measured in cellulose
```

This correction can be made as follows:
```{r}
d13C.disc.leaf <- (ATM.13C-(TR13-d))/(1+(TR13-d)/1000)
```

Leaf level ci can be calculated as follows:
```{r}
ci.leaf <-  ca*(d13C.disc.leaf-a)/(b-a)
```

Leaf level iwue can be calculated as follows:
```{r}
iwue.leaf <- (ca-ci.leaf)/1.6
```
now add the leaf level physiological indices to the previous plot:
```{r}
layout(matrix(1:6,nrow = 3, ncol = 2,byrow = F), widths = c(3,3,3),heights = c(3,3,3),respect = TRUE)
# Note you might need to adjust the ylim values.
ts.plot(TR13,ylab="d13C (permil.VPDB)",xlab="Years", main="OBSERVATIONS")
ts.plot(d13C.disc,ylab="D13C (permil.VPDB)",xlab="Years", main=""); lines(d13C.disc.leaf, col="#0B775E")
ts.plot(ca,ylab="CO2 (ppm)",xlab="Years", main="")
ts.plot(ci,ylab="ci (pa)",xlab="Years", main="FARQUHAR-MODEL"); lines(ci.leaf, col="#0B775E")
ts.plot(iwue,ylab=" iwue (umol/mol)",xlab="Years", main=""); lines(iwue.leaf, col="#0B775E")
ts.plot(ci/ca,ylab="ci/ca",xlab="Years", main=""); lines(ci.leaf/ca, col="#0B775E")
```
Question 4: What differences do you observe between physiological indices at stem level and leaf level?


The version of the Farquhar model we used above is a simplified description of the isotopic discrimination. It does not include the fractionation effects during the transfer of CO2 from substomatal cavities to the site of fixation via the mesophyll, and during mitochondrial respiration and photorespiration. While the mesophyll fractionation and their values remain highly unconstrained, recent studies have recommendedthe inclusion of the photorespiratory effect in the discrimination model as this term contributes to an increase of Δ13C with ca rise by 0.004‰ ppm-1 (Keeling et al., 2017; Lavergne et al., 2019).


A Farquhar model that includes the photorespiratory term is as follows:

Δ13C = a+(b-a)*(ci/ca) -f*(Gst/ca)

The photorespiratory term is the: f*(Gst/ca)

The fractionation due to photorespiration, f, is 12 permil. 

Gst (or Γ*, Gamma Star) is the CO2 compensation point in the absence of mitochondrial respiration (Pa), it is calculated from the temperature (T) and atmospheric pressure response.You can use the calc_gammastar(tc, patm) function from the rpmodel to calculate Gamma Star



Based on Notebook 1 and the temperature file, define the appropriate temperature window (e.g.average growing season). This has to be a Time series of site level temperature variability during the summer or growing season or relevant season for trees at this site. Below is an example for summer (June-July-August):
```{r}
gridded.tt.data <- ts(read.table("CRU_TT.txt",header=T),start=1901,frequency=1)
tt.JJA <- rowMeans(gridded.tt.data[,c(6,7,8)])
tt.JJA <- ts(tt.JJA ,start=1901,frequency =1)
```

Calculate gamma star as follows:
```{r}
tc <- tt.JJA # and patm has been defined earlier.
Gst <- calc_gammastar(tc, patm)
```

And the photorespiratory term as follows:
```{r}
photoresp_term <- 12*(Gst/ca) # ca is  in Pa
```
Calculate ci to include photorespiration:
```{r}
ci.leaf.photo <-  ca*(d13C.disc.leaf-a+photoresp_term)/(b-a)
```
And iwue to include photorespiration:
```{r}
iwue.leaf.photo <- (ca-ci.leaf.photo)/1.6
```

Plot the leaf level physiological parameters using the simple and photorespiration models:
```{r}

layout(matrix(1:6,nrow = 3, ncol = 2,byrow = F), widths = c(3,3,3),heights = c(3,3,3),respect = TRUE)
# Note you might need to adjust the ylim values.
ts.plot(TR13,ylab="d13C (permil.VPDB)",xlab="Years", main="OBSERVATIONS")
ts.plot(d13C.disc.leaf,ylab="D13C (permil.VPDB)",xlab="Years", main="")
ts.plot(ca,ylab="CO2 (pa)",xlab="Years", main="")
ts.plot(ci.leaf,ylab="ci (pa)",xlab="Years", main="FARQUHAR_MODEL"); lines(ci.leaf.photo, col="#0B775E")
ts.plot(iwue.leaf,ylab=" iwue (umol/mol)",xlab="Years", main=""); lines(iwue.leaf.photo, col="#0B775E")
ts.plot(ci.leaf/ca,ylab="ci/ca",xlab="Years", main=""); lines(ci.leaf.photo/ca, col="#0B775E")
legend("bottomleft",c("SIMPLE", "PHOTO"),lwd= c(1,1), bty = "n", col=c("black","#0B775E"),
       text.col=c("black","#0B775E"),ncol=1)

```

Question 5: Do the trends and amplitude of interannual variability vary between the simple model
and the one including photorespiration?

## PART 4: interpreting iWUE and leaf gas exchange strategies 

From the parts above, you have now calculated physiological parameters using best practices to account for most fractionation factors, using the most updated source data, and scaling to leaf level.

In the following, we are going to interpret variations of this physiological parameters. We will focus on those physiological parameters using the model with photorespiration.

First, compute the trends of iWUE, ci and ci/ca over the period of record. you can use the lm function and the summary to look at and report the statistics (R2, p value). Here is a first example with iWUE:
```{r}
iwue.linear <- lm(iwue.leaf.photo ~ seq(1901,2002,by=1))
summary(iwue.linear)
```
Repeat these steps for ci and ci/ca:
```{r}
# ci.linear <- 
```
Plot time-series of iWUE, ci and ci/ca and add the linear trend. Here is a first example with iWUE. You can use the layout to have 3 plots:
```{r}
layout(matrix(1:3,nrow = 1, ncol = 3,byrow = T), widths = c(3,3,3),heights = c(3,3,3),respect = TRUE)
ts.plot(iwue.leaf.photo,ylab=" iwue (umol/mol)",xlab="Years", main="") 
abline(iwue.linear, col="#0B775E",lwd=2,lty=3)

```

Question 6: Using computed statistics above, Describe the trends and their significance for each of the physiological indices.

Question 7: How do you interpret the ci/ca trends in term of stomatal conductance? 

Question 8: Was the iWUE increase steady over time?

Question 9: What was the rate of the ci (pa.year-1) increase over time? 

Question 10: Was this rate steady/constant over time?

Question 11: compared to the rate of ca, was the rate of ci: a) proportional   b) similar.To answer this question, you need to estimate the rate of ca using lm and summary functions.

Estimate the ci increase relative to ca. You can do this by regressing ci by ca.
```{r}
ci.linear <- lm(ci ~ ca)
summary(ci.linear)
```

Question 12: Report the rate and describe it in terms of a) proportional   b) similar. Note that this rate translate the ci increase for each ppm/pa of ca increase.

Question 13: Looking at all physiological indices, their trends and rates, how do you interpret the temporal variations of stomatal conductance and photosynthesis of these trees?


With rising ca, variations in plant Δ13C have been grouped into three leaf gas-exchange strategies:

(1) constant ci, (2) constant ci/ca, and (3) constant ca − ci. 

The first two strategies are considered active, and the third passive. These theoretical considerations serve as the basis for the interpretation of the physiological mechanisms underlying iWUE trends.

Calculate ci following scenario 1:
```{r}
# first, estimate the average ci from tree ring for the first decade of the record
ci.average_1decade <- mean(ci.leaf.photo[1:10])
ci.scenario1 <- ts(rep(ci.average_1decade, 102), start=1901, frequency=1)
```
Calculate ci following scenario 2:
```{r}
# first, estimate the average ci/ca from tree ring for the first decade of the record
cica.average_1decade <- mean((ci.leaf.photo/ca)[1:10])
ci.scenario2 <-ca*cica.average_1decade
```
Calculate ci following scenario 3:
```{r}
# first, estimate the average ca-ci from tree ring for the first decade of the record
caci.average_1decade <- mean((ca-ci.leaf.photo)[1:10])
ci.scenario3 <-ca-caci.average_1decade

```
Plot the observed ci with the predicted ci following the 3 scenarios:
```{r}
ts.plot(ci.leaf.photo,ylab="ci (pa)",xlab="Years", main="", ylim=c(15,25))
lines(ci.scenario1, lty=2, col="#F2AD00",lwd=2)
lines(ci.scenario2, lty=6, col="#F2AD00",lwd=2)
lines(ci.scenario3, lty=3, col="#F2AD00",lwd=2)
legend("topleft",c("ci-observed", "ci constant","ci/ca constant","ca-ci constant"),
       lty= c(1,2,6,3), lwd=c(1,2,2,2),bty = "n", col=c("black","#F2AD00","#F2AD00","#F2AD00"),
       text.col=c("black","#F2AD00","#F2AD00","#F2AD00"), ncol=1)
```

Question 14: Does the observed ci (derived from tree rings) follow any of the 3 scenarios?

Question 15:   Does the observed ci follow a scenario consistently throughout the record?

Compute iwue following the three scenarios (using ci for 3 scenarios from above) and plot with observed iwue:
```{r}
# iwue.scenario1 <- 
```

## Discussion 

From Notebook 1, you assessed climate correlations between 13C and climate variables. Based on the climate trends and tree ring d13C sensitivity to these climate factors/trends, answer the following:

Question 16: How do you interpret the change (increase) in iwue?

Question 17: is it driven by an increase is photosynthetic assimilation?

Question 18: is it driven by a decrease is stomatal conductance?

Question 19: are iwue, photosynthetic assimilation,stomatal conductance driven by

 a) rising ca, b) trends in climate, c) both?
 
Question 20: discuss how trends in ca and/or climate are affecting tree physiology
